---
title: "Data Store for Applicaiton"
subtitle: "Compiles data and metadata"
format: 
  html:
    embed-resources: true
    toc: true
editor: source
execute:
  warning: false
  message: false
editor_options: 
  chunk_output_type: console
---

```{r}
#| code-summary: "Setup"
#| code-fold: true

## instal pacman if required
if (!require(pacman)) install.packages("pacman")
pacman::p_load(
  tidyverse, arrow, here,
  reactable,
  sf, rmapshaper, leaflet, assertr,
  bslib, shiny
) 
setwd(here("App/shiny"))

```

Here we just do a initial EDA on the data before we do any shiny apps. It would also be very helpful if Tamara/Amber created notebooks with the visualizations they want to turn into a Shiny app. 



# Import 

So got an message that the final datasets are in `/clean datasets/council_measures_reduc.RData`; let's take a look!

## Data 

### Districts

```{r}
load(here("Data/clean datasets/council_measures_reduc.RData"))
df_data_raw = council_measures_reduc %>% 
  janitor::clean_names()
df_data_raw %>% glimpse()
```

Looks okay. Note that there is another data object that contains intermediate variables as well. Which can be loaded below.  For Now let's focus on the curated object.

```{{r}}
load(here("Data/clean datasets/council_measures_reduc_geom.RData"))
df_intermediate = council_measures_reduc_geom %>% 
  janitor::clean_names()
df_intermediate %>% glimpse()
```

Let's tidy

```{r}
df_data_int = df_data_raw %>% 
  sf::st_drop_geometry() %>% 
  pivot_longer(-district, names_to = 'var_name')


```

### City

The original data is not able to be mapped to the data we ahve. 

```{{r}}
load(here("Data/clean datasets/City_average.RData"))
df_city = City_average %>% 
  janitor::clean_names() %>% 
  mutate(id = 'city_avg') %>% 
  pivot_longer(-id, names_to = 'var_name') %>%
  rename(var_name_city_raw = var_name)
df_city %>%  count(var_name_city_raw)
```

Let's just do our own 

```{r}
df_pop_district = df_data_int %>% 
  filter(var_name == 'cc_district_population') %>% 
  select(district, pop_district = value)

df_city_avg = df_data_int %>% 
  filter(var_name != 'cc_district_population') %>%
  left_join(df_pop_district) %>% 
  summarise(city_avg = weighted.mean(value, pop_district),
            .by = 'var_name')
```

Let's append this to df_data and quarentine some variables from city average that doesn't make sense. 


```{r}
vec__remove_city_average = c(
  ''
)

df_data = df_data_int %>% 
  filter(var_name != 'cc_district_population') %>%
  left_join(df_city_avg) %>% 
  mutate(city_avg = ifelse(var_name  %in% vec__remove_city_average, NA, city_avg)) 
```

## Boundaries

Boundaries to City Districts are found in the original Rdata files processed by Amber. Let's just extract that and do a quicks sanity check. 

```{r}
path_amber_rdata_1 = list.files(path = here("Data/clean datasets"), 
           pattern = "\\Active_licenses.Rdata$", 
           full.names = TRUE)

load(path_amber_rdata_1)

sf_districts = Active_licenses %>% 
  janitor::clean_names() %>% 
  select(district)
glimpse(sf_districts)
  
```

Let's take a quick look.

```{r}
sf_districts %>% 
  leaflet() %>%
  addTiles() %>%
  addPolygons(
    fillColor = "red", 
    weight = 1, 
    fillOpacity = 0.2,
    # Add these lines for hover functionality
    highlightOptions = highlightOptions(
      weight = 3,
      color = "#666",
      fillOpacity = 0.7,
      bringToFront = TRUE
    ),
    # Create labels from your data
    label = ~paste("District:",  district),
    labelOptions = labelOptions(
      style = list("font-weight" = "normal", padding = "3px 8px"),
      textsize = "15px",
      direction = "auto"
    )
  )
```

Let's append the data and save it for the app. 

```{r}

library(geojsonio)
library(jsonlite)

geojson_districts <- sf_districts %>% 
  geojsonio::geojson_json() %>% 
  jsonlite::fromJSON(simplifyVector = FALSE)

geojson_districts_simp <- sf_districts %>% 
  rmapshaper::ms_simplify() %>%
  geojsonio::geojson_json() %>% 
  jsonlite::fromJSON(simplifyVector = FALSE)
```


## Metadata

Here we import the documentation specifically the `Docuemntaiton for Dashbaord` sheet from the documentation excel sheet.

```{r}
#| code-summary: "Shiny App UI"

library(readxl)
library(janitor)

df_metadata_raw =  read_excel(
  here("Data/data documentation/Data documentation for council variables.xlsx"),
  sheet = "Documentation for Dashboard") %>% 
  janitor::clean_names() 
df_metadata = df_metadata_raw %>% 
  rename(year_raw = year_s) %>% 
  rowwise() %>% 
  mutate(
       var_label = recode(
        metric_name,
        "Owners" = "Homeowners",
        "Uninsured" = "Without Health Insurance",
        "Black" = "Race and Ethnicity: Black",
        "White" = "Race and Ethnicity: non-Hispanic White",
        "Native American" = "Race and Ethnicity: Native American",
        "Asian" = "Race and Ethnicity: Asian",
        "Pacific Islander" = "Race and Ethnicity: Pacific Islander",
        "Other Race" = "Race and Ethnicity: Some other Race alone",
        "Two or More Races" = "Race and Ethnicity: Two or More Races",
        "Hispanic" = "Race and Ethnicity: Hispanic",
        "Less than High School" = "Education: Less than High School",
        "High School Graduate" = "Education: High School Graduate",
        "Some College" = "Education: Some College",
        "College Graduate" = "Education: College Graduate"
      ),
    year = case_when(
      data_source == 'ACS 5-year' & year_raw == "2018-2022" ~ "2022",
      data_source == 'Open Data Philly' & year_raw == "2020 - Oct 2024" ~ "2020-2024",
      data_source == 'Open Data Philly' & year_raw == "2019 - July 2024" ~ "2019-2024",
      data_source == 'Open Data Philly' & year_raw == "2015 - Present" ~ "2015-2024",
      data_source == 'Open Data Philly' & year_raw == "2019 - Present" ~ "2019-2024",
      TRUE ~ year_raw
    )
  ) %>% 
  ungroup() %>% 
  select(
    metric_name,
    var_label,
    var_def = metric_description_say_whether_it_is_a_percent_rate_count_etc,
    source = data_source,
    year,
    aggregation_notes = geography_of_original_estimate_and_aggregation_technique,
    cleaning_notes = notes
  ) %>% 
  mutate(
    var_def = recode(var_def,
                     "Percentage of people Uninsured" ="Percentage of people uninsured"),
    var_name = recode(
      metric_name,
      "Median Age" = "district_median_age_total",
      "Total population" = "cc_district_population",
      "Active Business Licenses No Rentals" = "total_active_licenses_norentals",
      "Active Business Licenses Rentals Only" = "total_active_licenses_rentalsonly",
      "Fatal Crashes" = "count_crashes",
      "Less Than High School" = "less_than_hs_pct",
      "High School Graduate" = "hs_grad_pct",
      "Some College" = "some_college_pct",
      "College Graduate" = "college_grad_pct",
      "Code Violations" = "pct_violations",
      "Lack Complete Kitchen" = "district_lack_kitch_pct",
      "Lack Complete Plumbing" = "district_lack_plumb_pct",
      "Median Household Income" = "median_hh_income_district",
      "Owners" = "pct_owner",
      "Renters" = "pct_renter",
      "White" = "pct_white",
      "Black" = "pct_black",
      "Native American" = "pct_native",
      "Asian" = "pct_asian",
      "Pacific Islander" = "pct_pi",
      "Other Race" = "pct_other",
      "Two or More Races" = "pct_two_more",
      "Hispanic" = "pct_hispanic",
      "Fatal Shootings" = "count_fatal",
      "Nonfatal Shootings" = "count_non_fatal",
      "Uninsured" = "percentage_uninsured",
      "Tree Canopy Coverage" = "cd_tree",
      "Heat Vulnerabillity Index" = "weighted_hvi",
      "Poverty" = "cd_pct_poverty"
    ),
    ylabs = case_when(
      str_detect(var_def, "Percentage") ~ paste(var_label, " (%)"),
      var_label %in% c('Median Age') ~ "Age",
      var_name %in% c('total_active_licenses_rentalsonly',
                      'total_active_licenses_norentals') ~ "# of Active Business Rentals",   
      var_def %in% c('Total Number of Fatal Shootings per Council District') ~ "# of Fatal Shootings",
      var_def %in% c('Total Number of Nonfatal Shootings per Council District') ~ "# of Nonfatal Shootings",
      var_label %in% c('Median Household Income') ~ "Median Household Income ($)",
      TRUE ~ var_def
    )
    ) %>% 
  filter(!var_name %in% c('cc_district_population')) %>% 
  ## validations
  assertr::verify(nchar(ylabs)<100)


## Check metadata-data compatibility
assertthat::assert_that(
  all(df_metadata$var_name %in% df_data$var_name),
  msg = "Metadata variable names do not match data variable names"
)
assertthat::assert_that(
  all(df_data$var_name %>% discard(~.x%in%c("district","geometry"))  %in% df_metadata$var_name),
  msg = "Metadata variable names do not match data variable names"
)

```

Let's add basic metadata to the data.

##  Cube

This is a key step to ensure interoperability between metadata and data:

1. Combine data and metadata
2. Operationalize hybrid metadata that utilizes both data/metadata

```{r}
library(scales)
df_data_cube = df_data %>% 
  left_join(df_metadata)  %>% 
  left_join(df_metadata_raw) %>% 
  mutate(
    district_int = as.integer(district),
    source_year =  paste0("Source: ", source," (",
                          year,")"),
    value_clean  = case_when(
      str_detect(var_def, "Percentage") ~ scales::percent(value/100, accuracy = 0.1),
      var_label %in% c('Median Age') ~ format(round(value, digits = 0), nsmall = 0),
      var_def %in% c('Total Number of Fatal Shootings per Council District',
                     'Total Number of Nonfatal Shootings per Council District',
                     'Count of crashes',
                     'Count of Licenses') ~ format(round(value, digits = 0), nsmall = 0) ,
      var_def %in% c('Heat Vulnerability Index') ~ format(round(value, digits = 2), nsmall = 2) ,
      var_label %in% c('Median Household Income') ~  scales::dollar(value, accuracy = 1),
      var_name %in% c('total_active_licenses_rentalsonly',
                      'total_active_licenses_norentals') ~ format(round(value, digits = 0), nsmall = 0),
      
      TRUE ~ "!!!!!!!!!!!!!!!!!!"
    ) %>% str_trim()
  ) %>%   
  ## Standardize
  arrange(var_name, district_int, year_s) %>%
  ## Unit testing
  verify(!str_detect(value_clean, "!!"))
```


# Export

## Zenodo Bundle

Connect to excell sheet.

```{r}

## Metadata
df_metadata_bundle = df_data_cube %>% 
  select(-district, -value, -city_avg,
         -district_int, -value_clean) %>% 
  distinct() %>% 
  arrange(var_name) %>% 
  ## Testing
  verify(nrow(.) == nrow(df_metadata))

df_data_bundle = df_data_cube %>% 
  select(district, metric_name, value)

```


Let's export

```{r}

df_metadata_bundle %>% 
  write_csv(here("Data/data documentation/download_bundles/v1.1/codebook_v1_1.csv"))

df_data_bundle %>% 
  write_csv(here("Data/data documentation/download_bundles/v1.1/data_v1_1.csv"))

```


## JSON


This is loading the data to the Javascript port of this R applciaiton. 


```{r}
library(tidyverse)
library(here)
library(arrow)
library(jsonlite)

load(here("App/shiny/phl-council-health-app-v1/data/app_v2.RData"))
geojson_districts = readRDS(here("App/shiny/phl-council-health-app-v1/data/geojson_districts_simp.rds"))

```


### df-metadata

This syncs with the application's typing whichis defined in `dashboard_types.ts`.

```{r}

df_metadata_bundle %>% 
  select(var_name, var_label, var_def, source, year, ylabs) %>% 
  write_json(here("App/shiny/next-js-deploy/public/data/df_metadata.json"), 
             auto_unbox = TRUE, 
             pretty = TRUE)
  
```


### df-data

This syncs with the application's typing whichis defined in `dashboard_types.ts`.

```{r}
df_data_cube %>% 
  select(district, var_name, value, city_avg, var_label, var_def, 
         source, year, aggregation_notes, cleaning_notes, ylabs, 
         district_int, source_year, value_clean) %>% 
  write_json(here("App/shiny/next-js-deploy/public/data/df_data.json"), 
             auto_unbox = TRUE)
```

### geojson-districts

```{r}
geojson_districts %>% 
  write_json(here("App/shiny/next-js-deploy/public/data/geojson_districts.json"), 
             auto_unbox = TRUE)
```