---
title: "phl-districts - 2025-03-25 - Initial data EDA"
format: 
  html:
    embed-resources: true
    toc: true
editor: source
execute:
  warning: false
  message: false
editor_options: 
  chunk_output_type: console
---

```{r}
#| code-summary: "Setup"
#| code-fold: true

## instal pacman if required
if (!require(pacman)) install.packages("pacman")
pacman::p_load(
  tidyverse, arrow, here,
  reactable,
  sf, rmapshaper, leaflet,
  bslib, shiny
) 
setwd(here("Data/notebooks/2025-03-25 - Initial data EDA"))

```

Here we just do a initial EDA on the data before we do any shiny apps. It would also be very helpful if Tamara/Amber created notebooks with the visualizations they want to turn into a Shiny app. 

# Import 

So got an message that the final datasets are in `/clean datasets/council_measures_reduc.RData`; let's take a look!

## Data 

### Districts

```{r}
load(here("Data/clean datasets/council_measures_reduc.RData"))
df_data_raw = council_measures_reduc %>% 
  janitor::clean_names()
df_data_raw %>% glimpse()
```

Looks okay. Note that there is another data object that contains intermediate variables as well. Which can be loaded below.  For Now let's focus on the curated object.

```{{r}}
load(here("Data/clean datasets/council_measures_reduc_geom.RData"))
df_intermediate = council_measures_reduc_geom %>% 
  janitor::clean_names()
df_intermediate %>% glimpse()
```

Let's tidy

```{r}
df_data_int = df_data_raw %>% 
  sf::st_drop_geometry() %>% 
  pivot_longer(-district, names_to = 'var_name')


```

### City

The original data is not able to be mapped to the data we ahve. 

```{{r}}
load(here("Data/clean datasets/City_average.RData"))
df_city = City_average %>% 
  janitor::clean_names() %>% 
  mutate(id = 'city_avg') %>% 
  pivot_longer(-id, names_to = 'var_name') %>%
  rename(var_name_city_raw = var_name)
df_city %>%  count(var_name_city_raw)
```

Let's just do our own 

```{r}
df_pop_district = df_data_int %>% 
  filter(var_name == 'cc_district_population') %>% 
  select(district, pop_district = value)

df_city_avg = df_data_int %>% 
  filter(var_name != 'cc_district_population') %>%
  left_join(df_pop_district) %>% 
  summarise(city_avg = weighted.mean(value, pop_district),
            .by = 'var_name')
```

Let's append this to df_data and quarentine some variables from city average that doesn't make sense. 


```{r}
vec__remove_city_average = c(
  ''
)

df_data_final = df_data_int %>% 
  filter(var_name != 'cc_district_population') %>%
  left_join(df_city_avg) %>% 
  mutate(city_avg = ifelse(var_name  %in% vec__remove_city_average, NA, city_avg)) 
```

## Boundaries

Boundaries to City Districts are found in the original Rdata files processed by Amber. Let's just extract that and do a quicks sanity check. 

```{r}
path_amber_rdata_1 = list.files(path = here("Data/clean datasets"), 
           pattern = "\\Active_licenses.Rdata$", 
           full.names = TRUE)

load(path_amber_rdata_1)

sf_districts = Active_licenses %>% 
  janitor::clean_names() %>% 
  select(district)
glimpse(sf_districts)
  
```

Let's take a quick look.

```{r}
sf_districts %>% 
  leaflet() %>%
  addTiles() %>%
  addPolygons(
    fillColor = "red", 
    weight = 1, 
    fillOpacity = 0.2,
    # Add these lines for hover functionality
    highlightOptions = highlightOptions(
      weight = 3,
      color = "#666",
      fillOpacity = 0.7,
      bringToFront = TRUE
    ),
    # Create labels from your data
    label = ~paste("District:",  district),
    labelOptions = labelOptions(
      style = list("font-weight" = "normal", padding = "3px 8px"),
      textsize = "15px",
      direction = "auto"
    )
  )
```

Let's append the data and save it for the app. 

```{r}

library(geojsonio)
library(jsonlite)

geojson_districts <- sf_districts %>% 
  geojsonio::geojson_json() %>% 
  jsonlite::fromJSON(simplifyVector = FALSE)
```


## Metadata

Here we import the documentation specifically the `Docuemntaiton for Dashbaord` sheet from the documentation excel sheet.

```{r}
#| code-summary: "Shiny App UI"

library(readxl)
library(janitor)

df_metadata_raw = df_metadata = read_excel(
  here("Data/data documentation/Data documentation for council variables.xlsx"),
  sheet = "Documentation for Dashboard") %>% 
  janitor::clean_names() 
df_metadata = df_metadata_raw %>% 
  rename(year_raw = year_s) %>% 
  rowwise() %>% 
  mutate(
    year = case_when(
      data_source == 'ACS 5-year' & year_raw == "2018-2022" ~ "2022",
      data_source == 'Open Data Philly' & year_raw == "2020 - Oct 2024" ~ "2020-2024",
      data_source == 'Open Data Philly' & year_raw == "2019 - July 2024" ~ "2019-2024",
      data_source == 'Open Data Philly' & year_raw == "2015 - Present" ~ "2015-2024",
      data_source == 'Open Data Philly' & year_raw == "2019 - Present" ~ "2019-2024",
      TRUE ~ year_raw
    )
  ) %>% 
  ungroup() %>% 
  select(
    var_label = metric_name,
    var_def = metric_description_say_whether_it_is_a_percent_rate_count_etc,
    source = data_source,
    year,
    aggregation_notes = geography_of_original_estimate_and_aggregation_technique,
    cleaning_notes = notes
  )  %>% 
  mutate(
    var_name = recode(
      var_label,
      "Median Age" = "district_median_age_total",
      "Total population" = "cc_district_population",
      "Active Business Licenses No Rentals" = "total_active_licenses_norentals",
      "Active Business Licenses Rentals Only" = "total_active_licenses_rentalsonly",
      "Fatal Crashes" = "count_crashes",
      "Less Than High School" = "less_than_hs_pct",
      "High School Graduate" = "hs_grad_pct",
      "Some College" = "some_college_pct",
      "College Graduate" = "college_grad_pct",
      "Code Violations" = "pct_violations",
      "Lack Complete Kitchen" = "district_lack_kitch_pct",
      "Lack Complete Plumbing" = "district_lack_plumb_pct",
      "Median Household Income" = "median_hh_income_district",
      "Owners" = "pct_owner",
      "Renters" = "pct_renter",
      "White" = "pct_white",
      "Black" = "pct_black",
      "Native American" = "pct_native",
      "Asian" = "pct_asian",
      "Pacific Islander" = "pct_pi",
      "Other Race" = "pct_other",
      "Two or More Races" = "pct_two_more",
      "Hispanic" = "pct_hispanic",
      "Fatal Shootings" = "count_fatal",
      "Nonfatal Shootings" = "count_non_fatal",
      "Uninsured" = "percentage_uninsured",
      "Tree Canopy Coverage" = "cd_tree",
      "Heat Vulnerabillity Index" = "weighted_hvi",
      "Poverty" = "cd_pct_poverty"
    )) %>% 
  filter(!var_name %in% c('cc_district_population'))

## Check metadata-data compatibility
assertthat::assert_that(
  all(df_metadata$var_name %in% df_data_final$var_name),
  msg = "Metadata variable names do not match data variable names"
)
assertthat::assert_that(
  all(df_data_final$var_name %>% discard(~.x%in%c("district","geometry"))  %in% df_metadata$var_name),
  msg = "Metadata variable names do not match data variable names"
)

```

Let's add basic metadat to the data.

```{r}
df_data = df_data_final %>% 
  left_join(df_metadata %>% select(var_name, var_label, var_def, source, year))
```


# Export


```{r}
save(df_data, df_metadata, geojson_districts,sf_districts,
     file = here("App/shiny/phl-council-health-app-v1/data/app_v1.RData"))
```


# EDA

The goal of this section is to mockup of visualizations for dashboard. 

## Univariate Distribution analysis

Let's select variable to analyze.

```{r}
input = list()
input$select_variable = 'cc_district_population'

```

Let's write the visualization code

```{r}
df_data  %>% 
  filter(var_name == input$select_variable) %>% 
    ggplot(aes(x = value  )) +
    geom_density(fill = "blue", alpha = 0.5) +
    labs(title = paste("Density plot of ", input$select_variable))
```

## Spatial Distribution

Let's select variable to analyze.

```{r}
input = list()
input$select_variable = 'cc_district_population'

```

Let's write the visualization code with leaflet.

```{r}
# Create a color palette based on the variable
pal <- colorNumeric(
  palette = "viridis",
  domain = df_data[[input$select_variable]]
)

# Create the leaflet map

sf_districts %>% 
  left_join(df_data %>% filter(var_name == input$select_variable) )%>%
  leaflet() %>%
  addTiles() %>%
  addPolygons(
    fillColor = ~pal(value), 
    weight = 1, 
    opacity = 1,
    color = "white",
    dashArray = "3",
    fillOpacity = 0.7,
    # Add these lines for hover functionality
    highlightOptions = highlightOptions(
      weight = 3,
      color = "#666",
      fillOpacity = 0.9,
      bringToFront = TRUE
    )
   
  ) %>% 
  # Add a legend
  addLegend(
    position = "bottomright",
    pal = pal,
    values = ~value,
    title = input$select_variable,
    opacity = 0.7)
```
